Normalization by Evaluation for \\ Intuitionistic Propositional Logic

### 2018-07-23 notes/nbeSum.tex
ITC: fixes; write context consequently as subscripts

The interpretation $\den{A \imp B}$ of implication will follow the
Brouwer-Heyting-Kolmogorov (BHK) interpretation of intuitionistic
logic:
\begin{center}\itshape
  A proof of $A \imp B$ is a method turning any proof of $A$ into a
  proof of $B$.
\end{center}
However, the direct lifting of the functions space $\den{A \imp
  B}_\Gamma = \den A_\Gamma \to \den B_\Gamma$ does not work.  It does
not give a presheaf because of the contravariant occurrence of
presheaf $\den A$.  Instead, we force monotonicity by quantifying over
all OPEs of $\Gamma$:
\[
  \den{A \imp B}_\Gamma = \prod_{\Delta : \Cxt} \prod_{\tau : \Delta \leq \Gamma} (\den A
  _\Delta \to \den B _\Delta)
\]
Reification is also challenging.  Let $f : \den{A \imp B}_\Gamma$ and
attempt
%$\tau : \Delta \leq \Gamma$
\[
  \reify[\Gamma]{A \imp B} f %\;\Delta\;\tau =
    = \impIP {\reify[\Gamma.A]B{(f\;(\weak[A]{\id[\Gamma]})}\;\mbox{?})}
\]
where hole ``?'' should be filled with an element of $\den B_{\Gamma.A}$
which intuitively should represent the new hypothesis $\top :
\Hyp[\Gamma.A]A$ introduced by $\timpI$.  Thus we need a natural
transformation $\Hyp A \todot \den A$ which \emph{reflects} variables
into the semantics.  In order to define it by
induction on $A$, we need to generalize it to
\[
  \treflect^A : \Ne A \todot \den A
  .
\]
Let us make sure we can define it at atoms and conjunctions:
\[
\begin{array}{lllll}
  \treflect^P & = & \tne & : & \Ne P \todot \Nf P \\
  \treflect^\top & = & !  & : & \Ne \top \todot \hatone \\
  \treflect^{A \land B} & = & \langle\, \treflect^A \comp \andE1  , \;
                             \treflect^B \comp \andE2 \rangle
    & : & \Ne{A \land B} \todot \den A \hattimes \den B \\
\end{array}
\]
Observe the eliminations $\andE i$ introduced to reflect conjunctive hypotheses!

Now, we can complete the story for implication.  Let $\tau : \Delta
\leq \Gamma$ and $a : \den A_\Delta$ and set:
\[
\begin{array}{lll}
  \treflect^{A \imp B} & : & \Ne{A \imp B} \todot \den{A \imp B} \\
  \reflect[\Gamma]{A \imp B} t\;\Delta\;\tau\;a & = &
    \reflect[\Delta]B{(\impEP {\Ne[\tau] B t} {(\reify[\Delta] A a)})}
\\[1ex]
  \treify^{A \imp B} & : & \den{A \imp B} \todot \Nf{A \imp B} \\
  \reify[\Gamma]{A \imp B} f & = &
    \impIP {\reify[\Gamma.A]B{(f\;(\weak[A]{\id[\Gamma]})}\;(\reflect[\Gamma.A]A{(\hyp\ttop)})}
\end{array}
\]

@@

Evaluation is turning an expression $t : \TmG A$ into a value in
$\den A$.  However, a direct definition $\teval^A : \Tm A \todot \den A$
by induction on the expression fails in case $\timpI$:
\[
\begin{array}{lll}
   \teval^{A \imp B}_\Gamma & : & \TmG A \to \den {A \imp B} _\Gamma \\
   \teval^{A \imp B}\,(\impI t)\;\Delta\;a & = & ? \\
\end{array}
\]
To be able to pass the argument $a$ to the function $\teval\,(\impI
t)$, we have to generalize evaluation to take an valuation of the
context into account.
\[
  \teval^{A \imp B}_\Gamma  : \den \Gamma \todot \den {A \imp B}
\]
Here, $\den \Gamma$ is the presheaf obtained as the product of the
hypotheses contained in $\Gamma$:
\[
\begin{array}{lll}
  \den{\cempty} & = & \hatone \\
  \den{\Gamma.A} & = & \den\Gamma \hattimes \den A \\
\end{array}
\]
Looking up values in the environment is the following family of
natural transformations:
\[
\begin{array}{l@{~}lcl}
  \multicolumn 2 l {\tlookup^A_\Gamma} & : & \HypG A \to (\den \Gamma \todot \den A) \\
  \lookup[\Gamma.A]A \ttop    & \Delta\,(\gamma,a) & = & a \\
  \lookup[\Gamma.B]A {\pop x} &\Delta\,(\gamma,a) & = & \lookup[\Gamma]A x\, \Delta\, \gamma \\
\end{array}
\]
\begin{exercise}
  Express $\tlookup$ via the projections
  $\pi_i : \A_1 \hattimes \A_2 \todot \A_i$
  out of the presheaf product.
\end{exercise}
We are ready to define evaluation:
\[
\begin{array}{lcl}
  \evalG A {\hyp x} & = & \lookupG A x \\
  \evalG {A \imp B} {\impI t}\,\Delta\,\gamma\,\Delta'\,(\tau : \Delta' \leq \Delta)\,a
    & = & \eval[\Gamma.A]B t\,\Delta'\,(\den\Gamma_\tau\gamma,a) \\
  \evalG {B} {\impE t u}\,\Delta\,\gamma & = &
    \evalG {A \imp B} t\,\Delta\,\gamma\,\Delta\,\id[\Delta]\,
      (\evalG A u\,\Delta\,\gamma) \\
\end{array}
\]
\begin{exercise}
  Complete the definition of $\teval$ for conjunctions.
\end{exercise}
\begin{exercise}
  Express $\teval$ in terms of the operations of the CCC of
  presheaves!
\end{exercise}

@@

Normalization is reification after evaluation in the identity
environment:
\[
\begin{array}{lcl}
\tnorm^A & : & \Tm A \todot \Nf A \\
\norm[\Gamma] A t & = & \reify[\Gamma]A {(\eval[\Gamma] A t \,\Gamma\, \env[\Gamma])}
\end{array}
\]
The identity environment $\env[\Gamma]$ is defined by induction on
$\Gamma$ as follows:
\[
\begin{array}{lcl}
  \env[\Gamma] & : & \den \Gamma _\Gamma \\
  \env[\cempty] & = & () \\
  \env[\Gamma.A] & = & (\den\Gamma_{\weak[A] \id[\Gamma]}(\env[\Gamma]), \;
                          \reflect[\Gamma.A] A {(\hyp \ttop)})
\end{array}
\]

@@

Given a presheaf $\A$, we define a new presheaf $\Cover \A$
inductively by the following constructors:
\begin{gather*}
  \ru{a : \A_\Gamma}{\returnC a : \Cover\A_\Gamma}
\qquad
  \ru{t : \NeG \bot}{\falseC t : \Cover\A_\Gamma}
\\[1ex]
  \ru{t : \NeG{C \lor D} \qquad
      c : \Cover\A_{\Gamma.C} \qquad
      d : \Cover\A_{\Gamma.D}
    }{\orC t c d : \Cover\A_\Gamma}
\end{gather*}
\begin{exercise}
  Show that $\Cover\A$ is indeed a presheaf.
\end{exercise}
We can think of $c : \Cover\A_\Gamma$ as a decision tree with leaves in $\A$
and nodes labeled by terms we case on.  Nodes of type $\tfalseC$ have
no subtrees since \emph{ex falsum quod libet}, and $\torC$ nodes
casing on $t : \NeG{C \lor D}$ have
two subtrees, one having a new hypothesis $C$, and one having a new
hypothesis $C$.

@@

Evaluation poses one more challenge. Consider the term:
\[
\ru{x : \HypG {C \lor D} \qquad
    f : \Tm[\Gamma.C]{A \imp B} \qquad
    g : \Tm[\Gamma.D.A]{B} \qquad
    a : \TmG A
  }{\impEP{\orE {(\hyp x)} f {(\impI g)}} a : \TmG B}
\]
There is a redex $\impEP {\impI g} a$ separated by a $\torE$.  The latter
would naturally evaluate to a case tree with functions in the
leaves.  To apply it to the evaluation of $a$, we have to apply each
leaf to that argument.  We facilitate this by implementing
\[
  \tpaste^{A \imp B} : \Cover{\den{A \imp B}} \todot \den{A \imp B}
\]
that turns a case tree of functions into a function, which expects an
argument that will then be passed to all the functions in the case
tree.  In fact, we need pasting at all formulas $A$:
\[
\begin{array}{lcl}
\tpaste^A & : & \tCover\,{\den A} \todot \den A \\[1ex]
\tpaste^P & = & \tpasteNf \\
\tpaste^\bot & = & \tjoinC \\
\tpaste^{A \lor B} & = & \tjoinC \\
\tpaste^{A \land B} & = & \langle\,
  \tpaste^A \comp \Cover{\pi_1} \, , \;
  \tpaste^B \comp \Cover{\pi_2} \,\rangle \\[1ex]
\multicolumn 3 {l} {\paste[\Gamma]{A \imp B} c \;
  \Delta \; (\tau : \Delta{\leq}\Gamma) \; (a : \den A _\Delta) \, = \,
  (\tpaste^B \comp \mapC \phi \comp \Cover{\den {A \imp B}_\tau}\, c)
}
\end{array}
\]
Here, we need a stronger version of functoriality of $\tCover$ which
only requires an
\[
  \phi : \prod_{\Phi} \prod_{\delta : \Phi \leq \Delta} \A_\Phi
             \to \B_\Phi
\]
to produce a $\tmapC_\Delta t\; \phi : \Cover\A_\Delta \to \Cover\B_\Delta$.
In our case, $\A = \den{A \imp B}$ and $\B = \den{B}$ and
\[
  \phi\;\Phi\;\delta\;f = f\;\id[\Phi]\;(\den A _\delta\, a)
  .
\]
Pasting allows us to complete the definition of evaluation.
\[
\begin{array}{lcl}
  \teval_\Gamma^A & : & \TmG A \to \den \Gamma \todot \den A \\[1ex]
  \evalG C {\orE t u v} \Delta \; \gamma & = &
    \tpaste^C \;
     (\mapC {(\lambda \Phi\, (\tau  : \Phi \leq \Delta) \to [f,g])}\;c) \\[1ex]
  \multicolumn 1 r {\mbox{ where } c} & = &
      \evalG{A \lor B} t\;\Delta\;\gamma \\
  \multicolumn 1 r {f\;a} & = &
    \eval[\Gamma.A] C u \;\Phi\;(\den\Gamma_\tau\,\gamma, a) \\
  \multicolumn 1 r {g\;b} & = &
    \eval[\Gamma.B] C v \;\Phi\;(\den\Gamma_\tau\,\gamma, b) \\
  \\
\end{array}
\]
\begin{exercise}
  Really complete the definition of $\teval$ (all missing cases).
\end{exercise}
That's it!

>>>>

The interpretation $\den{A \imp B}$ of implication will follow the
Brouwer-Heyting-Kolmogorov (BHK) interpretation of intuitionistic
logic:
\begin{center}\itshape
  A proof of $A \imp B$ is a method turning any proof of $A$ into a
  proof of $B$.
\end{center}
However, the direct lifting of the functions space $\den{A \imp
  B}_\Gamma = \den A_\Gamma \to \den B_\Gamma$ does not work.  It does
not give a presheaf because of the contravariant occurrence of
presheaf $\den A$.  Instead, we force monotonicity by quantifying over
all OPEs of $\Gamma$:
\[
  \den{A \imp B}_\Gamma = \prod_{\Delta : \Cxt} \prod_{\tau : \Delta \leq \Gamma} (\den A
  _\Delta \to \den B _\Delta)
\]
Reification is also challenging.  Let $f : \den{A \imp B}_\Gamma$ and
attempt
%$\tau : \Delta \leq \Gamma$
\[
  \reify[\Gamma]{A \imp B} f %\;\Delta\;\tau =
    = \impIP {\reify[\Gamma.A]B{(f_{\Gamma.A}\;(\weak[A]{\id[\Gamma]})}\;\mbox{?})}
\]
where hole ``?'' should be filled with an element of $\den B_{\Gamma.A}$
which intuitively should represent the new hypothesis $\top :
\Hyp[\Gamma.A]A$ introduced by $\timpI$.  Thus we need a natural
transformation $\Hyp A \todot \den A$ which \emph{reflects} variables
into the semantics.  In order to define it by
induction on $A$, we need to generalize it to
\[
  \treflect^A : \Ne A \todot \den A
  .
\]
Let us make sure we can define it at atoms and conjunctions:
\[
\begin{array}{lllll}
  \treflect^P & = & \tne & : & \Ne P \todot \Nf P \\
  \treflect^\top & = & !  & : & \Ne \top \todot \hatone \\
  \treflect^{A \land B} & = & \langle\, \treflect^A \comp \andE1  , \;
                             \treflect^B \comp \andE2 \rangle
    & : & \Ne{A \land B} \todot \den A \hattimes \den B \\
\end{array}
\]
Observe the eliminations $\andE i$ introduced to reflect conjunctive hypotheses!

Now, we can complete the story for implication.  Let $\tau : \Delta
\leq \Gamma$ and $a : \den A_\Delta$ and set:
\[
\begin{array}{lll}
  \treflect^{A \imp B} & : & \Ne{A \imp B} \todot \den{A \imp B} \\
  \reflect[\Gamma]{A \imp B} t\;\Delta\;\tau\;a & = &
    \reflect[\Delta]B{(\impEP {\Ne[\tau] B t} {(\reify[\Delta] A a)})}
\\[1ex]
  \treify^{A \imp B} & : & \den{A \imp B} \todot \Nf{A \imp B} \\
  \reify[\Gamma]{A \imp B} f & = &
    \impIP {\reify[\Gamma.A]B{(f_{\Gamma.A}\;(\weak[A]{\id[\Gamma]})}\;(\reflect[\Gamma.A]A{(\hyp\ttop)})}
\end{array}
\]

@@

Evaluation is turning an expression $t : \TmG A$ into a value in
$\den A$.  However, a direct definition $\teval^A : \Tm A \todot \den A$
by induction on the expression fails in case $\timpI$:
\[
\begin{array}{lcl}
   \teval^{A \imp B}_\Gamma & : & \TmG A \to \den {A \imp B} _\Gamma \\
   \teval^{A \imp B}\,(\impI t)\,_\Delta\,a & = & ? \\
\end{array}
\]
To be able to pass the argument $a$ to the function $\teval\,(\impI
t)$, we have to generalize evaluation to take an valuation of the
context into account.
\[
  \evalG{A \imp B}t  ~:~ \den \Gamma \todot \den {A \imp B}
\]
Here, $\den \Gamma$ is the presheaf obtained as the product of the
hypotheses contained in $\Gamma$:
\[
\begin{array}{lll}
  \den{\cempty} & = & \hatone \\
  \den{\Gamma.A} & = & \den\Gamma \hattimes \den A \\
\end{array}
\]
Looking up values in the environment is the following family of
natural transformations:
\[
\begin{array}{l@{~}lcl}
  \multicolumn 2 l {\tlookup^A_\Gamma} & : & \HypG A \to (\den \Gamma \todot \den A) \\
  \lookup[\Gamma.A]A \ttop    &_\Delta\,(\gamma,a) & = & a \\
  \lookup[\Gamma.B]A {\pop x} &_\Delta\,(\gamma,a) & = & \lookup[\Gamma]A x\,_\Delta\, \gamma \\
\end{array}
\]
\begin{exercise}
  Express $\tlookup$ via the projections
  $\pi_i : \A_1 \hattimes \A_2 \todot \A_i$
  out of the presheaf product.
\end{exercise}
We are ready to define evaluation:
\[
\begin{array}{lcl}
  \evalG A {\hyp x} & = & \lookupG A x \\
  \evalG {A \imp B} {\impI t}\,_\Delta\,\gamma\,_{\Delta'}\,(\tau : \Delta' \leq \Delta)\,a
    & = & \eval[\Gamma.A]B t\,_{\Delta'}\,(\den\Gamma_\tau\gamma,a) \\
  \evalG {B} {\impE t u}\,_\Delta\,\gamma & = &
    \evalG {A \imp B} t\,_\Delta\,\gamma\,_\Delta\,\id[\Delta]\,
      (\evalG A u\,_\Delta\,\gamma) \\
\end{array}
\]
\begin{exercise}
  Complete the definition of $\teval$ for conjunctions.
\end{exercise}
\begin{exercise}
  Express $\teval$ in terms of the operations of the CCC of
  presheaves!
\end{exercise}

@@

Normalization is reification after evaluation in the identity
environment:
\[
\begin{array}{lcl}
\tnorm^A & : & \Tm A \todot \Nf A \\
\norm[\Gamma] A t & = & \reify[\Gamma]A {(\eval[\Gamma] A t \,_\Gamma\, \env[\Gamma])}
\end{array}
\]
The identity environment $\env[\Gamma]$ is defined by induction on
$\Gamma$ as follows:
\[
\begin{array}{lcl}
  \env[\Gamma] & : & \den \Gamma _\Gamma \\
  \env[\cempty] & = & () \\
  \env[\Gamma.A] & = & (\den\Gamma_{\weak[A] \id[\Gamma]}(\env[\Gamma]), \;
                          \reflect[\Gamma.A] A {(\hyp \ttop)})
\end{array}
\]

@@

Given a presheaf $\A$, we define a new presheaf $\Cover \A$
inductively by the following constructors:
\begin{gather*}
  \ru{a : \A_\Gamma}{\returnC a : \Cover\A_\Gamma}
\qquad
  \ru{t : \NeG \bot}{\falseC t : \Cover\A_\Gamma}
\\[1ex]
  \ru{t : \NeG{C \lor D} \qquad
      c : \Cover\A_{\Gamma.C} \qquad
      d : \Cover\A_{\Gamma.D}
    }{\orC t c d : \Cover\A_\Gamma}
\end{gather*}
\begin{exercise}
  Show that $\Cover\A$ is indeed a presheaf.
\end{exercise}
We can think of $c : \Cover\A_\Gamma$ as a decision tree with leaves in $\A$
and nodes labeled by terms we case on.  Nodes of type $\tfalseC$ have
no subtrees since \emph{ex falsum quod libet}, and $\torC$ nodes
casing on $t : \NeG{C \lor D}$ have
two subtrees, one having a new hypothesis $C$, and one having a new
hypothesis $D$.

@@

Evaluation poses one more challenge. Consider the term:
\[
\ru{x : \HypG {C \lor D} \qquad
    f : \Tm[\Gamma.C]{A \imp B} \qquad
    g : \Tm[\Gamma.D.A]{B} \qquad
    a : \TmG A
  }{\impEP{\orE {(\hyp x)} f {(\impI g)}} a : \TmG B}
\]
There is a redex $\impEP {\impI g} a$ separated by an $\torE$.
The function term $\orE {(\hyp x)} f {(\impI g)}$
would naturally evaluate to a case tree with functions in the
leaves.  To apply it to the evaluation of $a$, we have to apply each
leaf to that argument.  We facilitate this by implementing
\[
  \tpaste^{A \imp B} : \Cover{\den{A \imp B}} \todot \den{A \imp B}
\]
that turns a case tree of functions into a function which expects an
argument that will then be passed to all the functions in the case
tree.  In fact, we need pasting at all formulas $A$:
\[
\begin{array}{lcl}
\tpaste^A & : & \tCover\,{\den A} \todot \den A \\[1ex]
\tpaste^P & = & \tpasteNf \\
\tpaste^\bot & = & \tjoinC \\
\tpaste^{A \lor B} & = & \tjoinC \\
\tpaste^{A \land B} & = & \langle\,
  \tpaste^A \comp \Cover{\pi_1} \, , \;
  \tpaste^B \comp \Cover{\pi_2} \,\rangle \\[1ex]
\multicolumn 3 {l} {\paste[\Gamma]{A \imp B} c \;
  _\Delta \; (\tau : \Delta{\leq}\Gamma) \; (a : \den A _\Delta) \, = \,
  (\paste[\Delta]B \comp \mapC[\Delta] \varphi \comp \Cover{\den {A \imp B}_\tau})\, c
}
\end{array}
\]
In the case of implication $A \imp B$,
we need a stronger version of functoriality of $\tCover$ which
only requires a
\[
  \varphi : \prod_{\Phi} \prod_{\delta : \Phi \leq \Delta} \A_\Phi
             \to \B_\Phi
\]
to produce a $\mapC[\Delta]\varphi : \Cover\A_\Delta \to \Cover\B_\Delta$.
In our case, $\A = \den{A \imp B}$ and $\B = \den{B}$ and
\[
  \varphi\,_\Phi\;\delta\;f = f\;\id[\Phi]\;(\den A _\delta\, a)
  .
\]
Pasting allows us to complete the definition of evaluation.
\[
\begin{array}{lcl}
  \teval_\Gamma^A & : & \TmG A \to \den \Gamma \todot \den A \\[1ex]
  \evalG C {\orE t u v} _\Delta \; \gamma & = &
    \paste[\Delta]C
     {(\mapC[\Delta]{\lambda\, _\Phi\, (\tau  : \Phi \leq \Delta) \to [f,g]}\;c)} \\[1ex]
  \multicolumn 1 r {\mbox{ where } c} & = &
      \evalG{A \lor B} t\;_\Delta\;\gamma \\
  \multicolumn 1 r {f\;a} & = &
    \eval[\Gamma.A] C u \;_\Phi\;(\den\Gamma_\tau\,\gamma, a) \\
  \multicolumn 1 r {g\;b} & = &
    \eval[\Gamma.B] C v \;_\Phi\;(\den\Gamma_\tau\,\gamma, b) \\
  \\
\end{array}
\]
\begin{exercise}
  Really complete the definition of $\teval$ (all missing cases).
\end{exercise}
That's it!
